Health and disease imprinted in the time variability of the human microbiome

### 2016-04-15 biorxiv/arxiv_sub.tex
Move docx to MSWord folder in mSystems and add biorxiv files
On branch master
	new file:   biorxiv/arxiv_sub.tex
	new file:   biorxiv/arxiv_suppl_sub.tex
	renamed:    mSystems/WHOLE_draft_0.1.docx -> mSystems/MSWord/WHOLE_draft_0.1.docx
	renamed:    mSystems/discussion_draft_0.1.docx -> mSystems/MSWord/discussion_draft_0.1.docx
	renamed:    mSystems/intro_draft_0.1.docx -> mSystems/MSWord/intro_draft_0.1.docx
	renamed:    mSystems/mat&met_draft_0.1.docx -> mSystems/MSWord/mat&met_draft_0.1.docx
	renamed:    mSystems/results_draft_0.1.docx -> mSystems/MSWord/results_draft_0.1.docx
	renamed:

% Use only LaTeX2e, calling the article.cls class and 12-point type.

\documentclass[12pt]{article}

% Users of the {thebibliography} environment or BibTeX should use the
% scicite.sty package, downloadable from *Science* at
% www.sciencemag.org/about/authors/prep/TeX_help/ .
% This package should properly format in-text
% reference calls and reference-list numbers.

\usepackage{scicite}
\usepackage{graphicx}

% Use times if you have the font installed; otherwise, comment out the
% following line.

\usepackage{times}

% The preamble here sets up a lot of new/revised commands and
% environments.  It's annoying, but please do *not* try to strip these
% out into a separate .sty file (which could lead to the loss of some
% information when we convert the file to other formats).  Instead, keep
% them in the preamble of your main LaTeX source file.

% The following parameters seem to provide a reasonable page setup.

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm

%The next command sets up an environment for the abstract to your paper.

\newenvironment{sciabstract}{%
\begin{quote} \bf}
{\end{quote}}

% If your reference list includes text notes as well as references,
% include the following line; otherwise, comment it out.

\renewcommand\refname{References and Notes}

% The following lines set up an environment for the last note in the
% reference list, which commonly includes acknowledgments of funding,
% help, etc.  It's intended for users of BibTeX or the {thebibliography}
% environment.  Users who are hand-coding their references at the end
% using a list environment such as {enumerate} can simply add another
% item at the end, and it will be numbered automatically.

\newcounter{lastnote}
\newenvironment{scilastnote}{%
\setcounter{lastnote}{\value{enumiv}}%
\addtocounter{lastnote}{+1}%
\begin{list}%
{\arabic{lastnote}.}
{\setlength{\leftmargin}{.22in}}
{\setlength{\labelsep}{.5em}}}
{\end{list}}

% Include your paper's title here

\title{Microbiota: are you sick?}

% Place the author information here.  Please hand-code the contact
% information and notecalls; do *not* use \footnote commands.  Let the
% author contact information appear immediately below the author names
% as shown.  We would also prefer that you don't change the type-size
% settings shown here.

\author
{Jose Manuel Mart\'i,$^{1}$ Daniel M. Mart\'inez,$^{2}$ Manuel Pe\~na$^{1}$, C\'esar Gracia$^{1}$, \\
Amparo Latorre$^{2,3,4}$, Andr\'es Moya$^{2,3,4}$ \& Carlos P. Garay$^{1,\ast}$\\
\\
\normalsize{$^{1}$Instituto de Fisica Corpuscular, CSIC-UVEG, P.O.  22085, 46071, Valencia, Spain.}\\
\normalsize{$^{2}$FISABIO, Avda de Catalunya, 21, 46020, Valencia, Spain.}\\
\normalsize{$^{3}$Cavanilles Institute of Biodiversity and Evolutionary Biology, Univ. de Valencia, 46980, Spain.}\\
\normalsize{$^{4}$CIBER en Epidemiolog\' ia y Salud P\'ublica (CIBEResp), Madrid, Spain.}\\
\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  penya@ific.uv.es}
}

% Include the date command, but leave its argument blank.

\date{}

%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%

@@

% Double-space the manuscript.

\baselineskip24pt

% Make the title.

@@

% Place your abstract within the special {sciabstract} environment.

\begin{sciabstract}
Human microbiota plays an important role in determining changes from health to disease. Increasing research activity is dedicated to understand its diversity and variability. 
We analyse 16S rRNA and whole genome sequencing (WGS) data from the gut microbiota of 97 individuals monitored in time. Temporal fluctuations in the microbiome reveal 
significant differences due to factors that affect the microbiota such as dietary changes, antibiotic intake, early gut development or disease. Here we show that a fluctuation 
scaling law describes the temporal variability of the system and that a noise-induced phase transition is central in the route to disease. The universal law distinguishes healthy 
from sick microbiota and quantitatively characterizes the path in the phase space, which opens up its potential clinical use and, more generally, other technological 
applications where microbiota plays an important role. 
\end{sciabstract}

% In setting up this template for *Science* papers, we've used both
% the \section* command and the \paragraph* command for topical
% divisions.  Which you use will of course depend on the type of paper
% you're writing.  Review Articles tend to have displayed headings, for
% which \section* is more appropriate; Research Articles, when they have
% formal topical divisions at all, tend to signal them with bold text
% that runs into the paragraph, for which \paragraph* is the right
% choice.  Either way, use the asterisk (*) modifier, as shown, to
% suppress numbering.

\section*{Introduction}

@@

\begin{figure}[!ht]
\includegraphics[width=1.05\textwidth]{comboplot1.pdf}
\caption{Taylor's law parameter space. X-weighted power-law fits of the standard deviations versus the mean values for each bacterial genus monitored in time. We show the fit for samples from a healthy subject (Figure A) and from a subject diagnosed with irritable bowel syndrome (Figure B) studied in our lab\cite{durban}. We have compiled all data studied in this work in Figure C. The coloured region (dashed line) corresponds to 68\% (95\%) CL region of healthy individuals in the Taylor parameter space. Points with errors place each individual gut microbiome in the Taylor space. Note that the parameters have been standardized (stdev units) to the healthy group in each study for demonstrative purposes.}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=1.05\textwidth]{comboplot2.pdf}
\caption{Phase space of microbiota stability. We show the Variability parameter as a function of time for the male in Caporaso's study\cite{moving} (Figure A) and for the patient D in the antibiotics study\cite{antibiotic} (Figure B). Periods of antibiotic treatment are shown by blue vertical lines. Microbiota states can be placed in the phase space F-V (Figure C). The light blue shaded region corresponds to the stable phase, while the grey shaded region is the unstable phase (the phase transition line is calculated for  $\alpha$ = $\beta$ = 0.75). We place healthy individuals (green) and individuals whose gut microbiota is threatened (antibiotics, IBS) in the phase space fitness - variability. Gut microbiota of healthy individuals over a long term span show a quasi--periodical variability (central period is ten days). We show that taking antibiotics (AB1 and AB2 correspond to first and second treatment respectively) induces a phase transition in the gut microbiota, which impacts its future changes. We also show an IBS--diagnosed patient transiting from the unstable to the stable phase.}
\end{figure}

@@

\begin{thebibliography}{1}
\bibitem{moving} Caporaso, J.G. et al. Moving pictures of the human microbiome. {\it Genome Biol.} {\bf 12,} R50 (2011).
\bibitem{lea} Faith, J.J. et al. The long-term stability of the human gut microbiota. {\it Science} {\bf 341,} 1237439 (2013).
\bibitem{kwashiorkor} Smith M.I. et al. Gut microbiomes of Malawian twin pairs discordant for kwashiorkor. {\it Science} {\bf 339,} 548-54 (2013).
\bibitem{diet} David, L.A. et al. Diet rapidly and reproducibly alters the human gut microbiome. {\it Nature} {\bf 505,} 559-63 (2014).
\bibitem{antibiotic} Dethlefsen L., Relman D. A. Incomplete recovery and individualized responses of the human distal gut microbiota to repeated antibiotic perturbation. {\it Proc. Nat. Acad. Sci. USA} {\bf 108,} 4554-61 (2011).
\bibitem{durban} Durban, A. \textit{et al.} Structural alterations of faecal and mucosa-associated bacterial communities in irritable bowel syndrome. {\it FEMS Microbiol. Ecol.} {\bf 86}, 581-9 (2013).
\bibitem{qiime} Caporaso, J.G. et al. QIIME allows analysis of high-throughput community sequencing data. {\it Nature Methods} {\bf 7,} 335-6 (2010).
\bibitem{lmat} Ames, S.K. et al. Scalable metagenomic taxonomy classification usng a reference genome database. {\it Bioinformatics} {\bf 29,} 2253-2260 (2013).
\bibitem{fs} Eisler,Z., Bartos,I., Kertesz,J. Fluctuation scaling in complex systems: Taylor's law and beyond. {\it Adv. Phys.} {\bf 57,} 85 (2008).
\bibitem{taylor} Taylor, L.R. Aggregation, Variance and the mean. {\it Nature} {\bf 189,} 732-35 (1961).
\bibitem{stat} Jorgensen,B., Martinez,J.R., Tsao,M. Asymptotic behaviour of the variance function. {\it Scand. J. Statist..} {\bf 21,} 223-243 (1994).
\bibitem{convergence1} Fronczak,A., Fronczak,P. Origins of Taylor's power law for fluctuation scaling in complex systems. {\it Phys. Rev. E} {\bf 81,} 066112 (2010).
\bibitem{convergence2} Kendal, W.S., Jorgensen,B. Taylor's power law and fluctuation scaling explained by a central-limit-like convergence. {\it Phys. Rev. E} {\bf 83,} 066115 (2011).
\bibitem{convergence3} Kendal, W.S., Jorgensen,B. Tweedie convergence: A mathematical basis for Taylor's power law. {\it Phys. Rev. E} {\bf 84,} 066120 (2011).
\bibitem{ranking} Blumm, N. et al. Dynamics of ranking processes in complex systems. {\it Phys. Rev. Lett.} {\bf 109,} 128701 (2012).
\end{thebibliography}

>>>>

\documentclass[aps,prl,twocolumn,superscriptaddress,showpacs]{revtex4}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage{bm}

@@

\title{Health and disease imprinted in the time variability of the human microbiome}
\author{Jose Manuel Mart\' i}
\affiliation{Instituto de Fisica Corpuscular, CSIC-UVEG, P.O.  22085, 46071, Valencia, Spain}
\author{Daniel M. Mart\' inez}
\affiliation{Instituto de Fisica Corpuscular, CSIC-UVEG, P.O.  22085, 46071, Valencia, Spain}
\affiliation{FISABIO, Avda de Catalunya, 21, 46020, Valencia, Spain}
\author{C\'esar Gracia}
\affiliation{Instituto de Fisica Corpuscular, CSIC-UVEG, P.O.  22085, 46071, Valencia, Spain}
\author{Amparo Latorre}
\affiliation{FISABIO, Avda de Catalunya, 21, 46020, Valencia, Spain}
\affiliation{Cavanilles Institute of Biodiversity and Evolutionary Biology, Univ. de Valencia, 46980, Spain}
\affiliation{CIBER en Epidemiolog\' ia y Salud P\'ublica (CIBEResp), Madrid, Spain}
\author{Andr\'es Moya}
\affiliation{FISABIO, Avda de Catalunya, 21, 46020, Valencia, Spain}
\affiliation{Cavanilles Institute of Biodiversity and Evolutionary Biology, Univ. de Valencia, 46980, Spain}
\affiliation{CIBER en Epidemiolog\' ia y Salud P\'ublica (CIBEResp), Madrid, Spain}
\author{Carlos P. Garay}
\affiliation{Instituto de Fisica Corpuscular, CSIC-UVEG, P.O.  22085, 46071, Valencia, Spain}

 \begin{abstract}
Human microbiota plays an important role in determining changes from health to disease. Increasing research activity is dedicated to understand its diversity and variability. 
We analyse 16S rRNA and whole genome sequencing (WGS) data from the gut microbiota of 97 individuals monitored in time. Temporal fluctuations in the microbiome reveal 
significant differences due to factors that affect the microbiota such as dietary changes, antibiotic intake, early gut development or disease. Here we show that a fluctuation 
scaling law describes the temporal variability of the system and that a noise-induced phase transition is central in the route to disease. The universal law distinguishes healthy 
from sick microbiota and quantitatively characterizes the path in the phase space, which opens up its potential clinical use and, more generally, other technological 
applications where microbiota plays an important role. 
 \end{abstract}

@@

@@

\begin{figure*}
\includegraphics[width=14cm,angle=0]{comboplot1.pdf}
\caption{Taylor's law parameter space. X-weighted power-law fits of the standard deviations versus the mean values for each bacterial genus monitored in time. We show the fit for samples from a healthy subject (Figure A) and from a subject diagnosed with irritable bowel syndrome (Figure B) studied in our lab\cite{durban}. We have compiled all data studied in this work in Figure C. The coloured region (dashed line) corresponds to 68\% (95\%) CL region of healthy individuals in the Taylor parameter space. Points with errors place each individual gut microbiome in the Taylor space. Note that the parameters have been standardized (stdev units) to the healthy group in each study for demonstrative purposes.}
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=14cm,angle=0]{comboplot2.pdf}
\caption{Phase space of microbiota stability. We show the Variability parameter as a function of time for the male in Caporaso's study\cite{moving} (Figure A) and for the patient D in the antibiotics study\cite{antibiotic} (Figure B). Periods of antibiotic treatment are shown by blue vertical lines. Microbiota states can be placed in the phase space F-V (Figure C). The light blue shaded region corresponds to the stable phase, while the grey shaded region is the unstable phase (the phase transition line is calculated for  $\alpha$ = $\beta$ = 0.75). We place healthy individuals (green) and individuals whose gut microbiota is threatened (antibiotics, IBS) in the phase space fitness - variability. Gut microbiota of healthy individuals over a long term span show a quasi--periodical variability (central period is ten days). We show that taking antibiotics (AB1 and AB2 correspond to first and second treatment respectively) induces a phase transition in the gut microbiota, which impacts its future changes. We also show an IBS--diagnosed patient transiting from the unstable to the stable phase.}
\end{figure*}

@@

\textit{Acknowledgements.---} This work is supported by Generalitat Valencia Prometeo Grants II/2014/050, II/2014/065, by the Spanish Grants FPA2011-29678 , BFU2012-39816-C02-01 of MINECO and by PITN- GA-2011-289442-INVISIBLES. JMM and DMM acknowledge FPI and FISABIO fellowships.

\begin{thebibliography}{99}

\bibitem{moving} Caporaso, J.G. et al. Moving pictures of the human microbiome. {\it Genome Biol.} {\bf 12,} R50 (2011).
\bibitem{lea} Faith, J.J. et al. The long-term stability of the human gut microbiota. {\it Science} {\bf 341,} 1237439 (2013).
\bibitem{kwashiorkor} Smith M.I. et al. Gut microbiomes of Malawian twin pairs discordant for kwashiorkor. {\it Science} {\bf 339,} 548-54 (2013).
\bibitem{diet} David, L.A. et al. Diet rapidly and reproducibly alters the human gut microbiome. {\it Nature} {\bf 505,} 559-63 (2014).
\bibitem{antibiotic} Dethlefsen L., Relman D. A. Incomplete recovery and individualized responses of the human distal gut microbiota to repeated antibiotic perturbation. {\it Proc. Nat. Acad. Sci. USA} {\bf 108,} 4554-61 (2011).
\bibitem{durban} Durban, A. \textit{et al.} Structural alterations of faecal and mucosa-associated bacterial communities in irritable bowel syndrome. {\it FEMS Microbiol. Ecol.} {\bf 86}, 581-9 (2013).
\bibitem{qiime} Caporaso, J.G. et al. QIIME allows analysis of high-throughput community sequencing data. {\it Nature Methods} {\bf 7,} 335-6 (2010).
\bibitem{lmat} Ames, S.K. et al. Scalable metagenomic taxonomy classification usng a reference genome database. {\it Bioinformatics} {\bf 29,} 2253-2260 (2013).
\bibitem{fs} Eisler,Z., Bartos,I., Kertesz,J. Fluctuation scaling in complex systems: Taylor's law and beyond. {\it Adv. Phys.} {\bf 57,} 85 (2008).
\bibitem{taylor} Taylor, L.R. Aggregation, Variance and the mean. {\it Nature} {\bf 189,} 732-35 (1961).
\bibitem{stat} Jorgensen,B., Martinez,J.R., Tsao,M. Asymptotic behaviour of the variance function. {\it Scand. J. Statist..} {\bf 21,} 223-243 (1994).
\bibitem{convergence1} Fronczak,A., Fronczak,P. Origins of Taylor's power law for fluctuation scaling in complex systems. {\it Phys. Rev. E} {\bf 81,} 066112 (2010).
\bibitem{convergence2} Kendal, W.S., Jorgensen,B. Taylor's power law and fluctuation scaling explained by a central-limit-like convergence. {\it Phys. Rev. E} {\bf 83,} 066115 (2011).
\bibitem{convergence3} Kendal, W.S., Jorgensen,B. Tweedie convergence: A mathematical basis for Taylor's power law. {\it Phys. Rev. E} {\bf 84,} 066120 (2011).
\bibitem{ranking} Blumm, N. et al. Dynamics of ranking processes in complex systems. {\it Phys. Rev. Lett.} {\bf 109,} 128701 (2012).
\end{thebibliography}
